### 2022-09-08 AISTATS2023/paper.tex
continued method description

\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathcal{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

\begin{itemize}
  \item We introduce Uncertainty DALE (UDALE), an extension of DALE that quantifies the uncertainty of the global explanation, i.e.~the level of heterogeneous effects hidden behind the global explanation.\item We provide an algorithm that automatically computes the optimal bin splitting for robustly estimating the explanatory quantities, i.e., the global effect and the uncertainty. 
  \item We formally prove that our method finds the optimal grouping of samples, minimizing the added uncertainty over the unavoidable heterogeneity that is the lower-bound of the objective.
  \item We provide empirical evaluation of the method in artificial and real datasets.
\end{itemize}

@@

We refer to random variables (rv) using uppercase \( X \), whereas to
simple variables with plain lowercase \( x \). Bold denotes a vector;
\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
interest \(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbb{x}_c)\), but we clarify that it corresponds to the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f_{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method. \footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} PDPlots formulate the feature
effect of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f_{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate
it as an expectation over the conditional
\(\mathcal{X}_c|\mathcal{X}_s\), i.e.,
\(f_{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathcal{X}_c|\mathcal{X}_s = x_s}[f(x_s,
\mathcal{X}_c)]\). ALE computes the global effect at \(x_s\) as an
integration of local effects. The local effects are measured as the
expected change on the output
\( \frac{\partial f(x_s, \mathcal{X}_c)}{\partial x_s} \) over the
conditional distribution \( \Xcb|\mathcal{X}_s\). The formula that
defines ALE is presented below:

\begin{equation}
  \label{eq:ALE}
  f_{\mathtt{ALE}}(x_s) = c + \int_{-\infty}^{x_s} \mathbb{E}_{\Xcb|\mathcal{X}_s=z}\left[\frac{\partial f(z, \mathcal{X}_c)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDPlots
integrate over unrealistic instances due to the use of the marginal
distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
resolve this issue using the conditional distribution
\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
effects. ALE plots resolve, both they have two drawbacks TODO add more
info.

@@

@@

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
function \(f(\cdot)\) is a deterministic predictor, the local effect
is also a deterministic quantity. Global methods summarize the local
explanations into a single quantity. Therefore, ALE summarizes all the
local explanations at \(x_s\) by averaging the local effects accross
all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
words, ALE globalizes the local explanations at \(x_s\) through the
expected change in the output \(y\) wrt \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
\end{equation}
%
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the model the uncertainty of
the effect at \(x_s\) as the variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
\end{equation}

\noindent
%
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the black-box function. It is important to clarify
that the variance is only a way to model the uncertainty. Other
statistical properties can also be used. Elaborate.

ALE computes the final effect at \(x_s\) by accumulating/integrating
the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE}). The
choice for \(x_{min}\) is not important, as it only affects the
centering of ALE plot along the vertical axis. Equivalently, we define
the accumulated uncertainty until the point \(x_s\), as the integral
of the variance of local effects:

\begin{equation}
  \label{eq:3}
  f_{\mathtt{ALE}}^{\sigma^2}(x_s) = \int_{z_1, min}^{x_s} \sigma^2(z) \partial z
\end{equation}

In ALE plots the absolute value at a specific point is not an
interpretable quantity (add ref to paper). The meaningful
interpretation is the effect at a specific point, i.e. what happens to
the output given a small change in the feature of interest. The
integration takes place only for making the visual interpretation
smooth, for example, to spot larger intervals where the effect of a
\textbf{local} change is continuously negative. The same stands for
the accumulated uncertainty. It is meaningful to measure, only, the
uncertainty at at a specific point to understand the heterogeneous
effects behind the averaged effect. In the experimental Section, we
propose appropriate visualizations to concentrate the user to this
effect. However, the accumulated uncertainty, i.e. the aggregate
uncertainty allong all the points of the plot, will help us define a
meaningfull objective for the interval spliting step. (TODO: check
here if the idea about whether the accumulated uncertainty is a good
metric for modeling the interaction strength)

\subsection{Uncertainty Quantification at an interval}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and on estimations based on the
limited instances of the training set. Therefore, it is impossible to
estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the
granularity of a point \(x_s\)\, since the possibility to observe a
sample in the interval \([x_s - h, x_s + h]\) is zero, when
\(h \to 0\). Therefore, for estimation purposes, we are obliged to
create larger intervals (\(h>0\)). The mean effect at the interval
\([z_1, z_2)\) is the mean of the expected effects:

@@

\noindent
Accordingly, the accumulated uncertainty (variance) at the interval
\([z_1, z_2)\) is:

@@

\noindent

\paragraph{Theorem 1.}

The accumulated variance at the interval \([z_1, z_2)\) is the
accumulation of the all variances inside the interval plus the
accumulated squared residuals:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
where \(\rho(z) = \mu(z) - \mu(z_1, z_2)\). The proof is at the
Appendix. Theorem 1 decouples the (observable) variance in the
interval \([z_1, z_2)\) into two terms. The first term
\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\) accumulates the variance
of the local explanations (Eq.~(\ref{eq:ALE_var})) for all points
inside the interval. The second term accumulates the squared
differences between the mean effect at the interval and mean of the
local explanations for all points inside the interval. The first term
accumulates the the uncertainty due to the natural characteristics of
the experiment, whereas, the second term adds extra uncertainty due to
limiting the granularity of the effect at the bin level.

\subsection{Estimation of Effect at an Interval}

@@

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}_k(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
  \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
\end{equation}

The approximation is unbiased only if the points are uniformly
distributed in \([z_{k-1}, z_k]\). (TODOs: Check what happens
otherwise).

@@

@@

\subsubsection{Algorithms}

>>>>

\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathbf{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

\begin{itemize}
\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
  quantifies the uncertainty of the global explanation, i.e.~the level
  of heterogeneous effects hidden behind the global explanation.
\item
  We provide an algorithm that automatically computes the optimal bin
  splitting for robustly estimating the explanatory quantities, i.e.,
  the global effect and the uncertainty.
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity that is the lower-bound of the objective.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

@@

We refer to random variables (rv) using uppercase \( X \), whereas to
simple variables with plain lowercase \( x \). Bold denotes a vector;
\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
interest \(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} PDPlots formulate the feature
effect of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
integration of local effects. The local effects are measured as the
expected change on the output
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
conditional distribution \( \Xcb|X_s\). The formula that
defines ALE is presented below:

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDPlots
integrate over unrealistic instances due to the use of the marginal
distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
resolve this issue using the conditional distribution
\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
effects. ALE plots resolve, both they have two drawbacks TODO add more
info.

@@

The terms \textit{uncertainty of explanation} and \textit{level of
  heterogeneous effects} refer to the same effect

@@

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
function \(f(\cdot)\) is a deterministic predictor, the local effect
is also a deterministic quantity. Global methods summarize the local
explanations into a single quantity. Therefore, ALE summarizes all the
local explanations at \(x_s\) by averaging the local effects across
all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
words, ALE globalizes the local explanations at \(x_s\), computing the
expected effect at \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the uncertainty of the local
effects at \(x_s\) as the variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the black-box function. It is important to clarify
that the variance is only a way to model the uncertainty. Other
statistical properties can also be used. Elaborate.

ALE computes the final effect at \(x_s\) by accumulating/integrating
the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE_accumulated_mean}). The
choice for \(x_{min}\) is not important, as it only affects the
centering of ALE plot along the vertical axis. Equivalently, we define
the accumulated uncertainty (variance) until the point \(x_s\), as the
integral of the variance of local effects:

\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent
In ALE plots the accumulated effect as absolute value is not an
interpretable quantity and is only used for better visual
interpretation\citep{Gromping2020MAEP}. The meaningful interpretation
is the effect at a specific point, i.e., what happens to the output
given a small change in the feature of interest. The integration takes
place only for making the visual interpretation easier, for example,
to spot larger intervals where the effect of a \textbf{local} change
is continuously constant. The same stands for the accumulated
uncertainty. The uncertainty has a meaninful interpretation as the
level of heterogeneous effects, only at a specific point. In the
experimental Section, we propose appropriate visualizations to
communicate to this effect to the user. However, the accumulated
uncertainty, i.e., the aggregated variance of the local effects, helps
us define a sensible objective for the interval spliting step. (TODO:
check here if the idea about whether the accumulated uncertainty is a
good metric for modeling the interaction strength)

\subsection{Uncertainty Quantification and Estimation at an Interval}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
the limited instances of the training set. Unfortunately, it is
impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at
the granularity of a point \(x_s\), because The possibility to observe
a sample in the interval \([x_s - h, x_s + h]\) is zero, when
\(h \to 0\). Therefore, we are obliged split the axis \(x_s\) into a
sequence of larger non-overlaping intervals and estimate mean and
variance at the interval resolution. The mean effect at an interval
\([z_1, z_2)\) is the mean of the expected effects:

@@

\noindent
Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
is:

@@

@@

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
\end{equation}

The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). (TODOs: Check what happens
otherwise).

@@

ALE authors\citep{apley2020visualizing} proposed an approximation of
(\ref{eq:ALE_accumulated_mean}) by partitioning the axis of the
feature of interest into a sequence of non-overlaping bin and
computing a single effect in each bin. Later, (TODO: add citation to
us), proposed the DALE (Differential ALE) approximation that decouples
the two steps. The local effects are computed once using the partial
derivatives on the available instances of the training set,
i.e. \(\frac{\partial f}{\partial x_s}\), and the bin-spliting step,
simply allocates the precomputed local effects into bins, for
estimating the expected effect in each bin. The formula that estimates
the accumulate effects is:

\begin{equation}
  f^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x}
  \frac{1}{|\mathcal{S}_k|} \sum_{i:\xb^i \in \mathcal{S}_k}
  [f_s(\xb^i)] = \Delta x \sum_{k=1}^{k_x} \hat{\mu}_k
 \label{eq:DALE}
\end{equation}

We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). Both
methods face the limitation that the partitioning into non-overlaping
intervals is done blindly. The user pass the total number of bins
\(K\) as a hyperparameter, the bins are defined with equal-size
spliting, and the training instances are allocated accordingly. This
approach is vulnerable to non-robust estimations. The mean effect is
often poorly approximated from a very small number of samples and the
mean effect of empty bins is interpolated from their
neighboors. Furthermore, in our case, we need sufficient sample
populations for estimating the variance of the approximation, apart
from the mean effect.

For overcoming this limitations, we reformulate the partitioning as a
clustering of the training instances into a sequence variable-size
intervals. The objective of the clustering problem is inspired

ALE requires splittingthe estimation of the

In this section, we introduce a framework

\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
at an interval \([z_1, z_2)\) is the accumulation of the all variances
plus the accumulation of squared residuals inside the interval:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is at the Appendix. Theorem 1 decouples the accumulated
variance at an interval, the only quantity we can estimate, into two
terms. The first term \(\int_{z_1}^{z_2} \sigma^2(z) \partial z\),
quantifies the uncertainty due to the natural characteristics of the
experiment and the second term adds extra uncertainty due to the
limited resolution.

@@

\subsection{Visualization of ALE with Uncertainty}
